Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-13 write-ups/PLDI/paper.tex
More intro shortening.

Irregular algorithms usually require \emph{blocking synchronization}
between workgroups, to balance load and communicate intermediate
results.  Many graph algorithms employ a level-by-level strategy, requiring a global barrier between each level.
Work stealing algorithms require each workgroup
to maintain a queue, which is typically protected by a mutex to enable
stealing by other workgroups.

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling, which in the context of GPUs means fair scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if workgroup $A$ holds a mutex, an unfair scheduler may cause
another workgroup $B$ to spin-wait forever for the mutex to be
released.  Similarly, an unfair scheduler can cause a workgroup to spin-wait
indefinitely at a global barrier so that other workgroups do not make progress
towards the barrier.

\paragraph{A degree of fairness: occupancy-bound execution}
GPU schedulers are \emph{not} fair. the current GPU
programming models---OpenCL~\cite{opencl2Spec}, \ADComment{Tyler, please add best CUDA since I've asked you to add it below in any case} CUDA~\cite{...}  and
HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
workgroups, and current GPUs do not
provide fair scheduling in practice.  Roughly speaking, each workgroup
executing a GPU kernel is mapped to a hardware \emph{compute
  unit}.\footnote{In practice the situation can be more complex: if a
  kernel is sufficiently simple and if workgroups are small enough,
  multiple workgroups may be mapped to a single compute unit.  We assume, for the purpose of the current discussion, that the kernel is sufficiently complex that this does not happen.}
%
If a kernel is executed with more workgroups than there are compute
units then at least two workgroups must share a single compute unit.
The simplest way for a GPU driver to handle this is via an \emph{occupancy-bound}
execution model~\cite{DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{Tyler, please add best cite for OB model; I guess our OOPSLA paper introduced the term, but we should cite that + Owens?}  With this model, once a workgroup has commenced execution on a
compute unit (it has become \emph{occupant}), the workgroup has
exclusive access to the compute unit until it has finished execution,
after which a subsequent workgroup can be bound to the compute unit.
Experiments suggest that the occupancy-bound execution model is widely
implemented by today's GPU devices and drivers~\cite{DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{For the previous cite, should we cite other papers as well as ours?}  It is
evidently a simple model for device vendors to implement efficiently.

The occupancy-bound execution model does not guarantee fair scheduling
between workgroups: if all compute units are occupied then a
not-yet-occupant workgroup will not be scheduled until some occupant
workgroup completes execution.  Yet the execution model \emph{does}
provide fair scheduling between \emph{occupant} workgroups, because
they are bound to separate compute units that operate in parallel.
Current implementations of irregular data parallel algorithms on GPUs
assume that the target GPU device provides the occupancy-bound
execution model.  They exploit the model by scheduling no more workgroups
than there are available compute units~\cite{...}. \ADComment{Tyler, please pick best cite for previous claim.} This works
because \emph{today}'s GPUs seem to provide the occupancy-bound
execution model.

@@

Despite its practical prevalence, none of the current GPU programming
models actually mandate occupancy-bound execution.  Further, there are
clear reasons why this model is undesirable from the point-of-view of
GPU vendors and end users.  First, the execution model does not enable
multi-tasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  If a long-running
kernel occupies all compute units, the GPU cannot be used for other
tasks, such as graphics rendering, in the meantime.  Precisely because
of this problem, operating systems employ a GPU \emph{watchdog} to
heuristically detect and kill long-running computations on the GPU
being used for display rendering~\cite{DBLP:conf/iwocl/SorensenD16}.  Second, \emph{energy throttling} is
an important concern for battery-powered devices, such as smartphones
and tablets~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to be able to
power down one or more compute units if the battery level is low or if
the user puts the device into a low power state.  Implementing energy
throttling would require being able to suspend the execution of a
workgroup, power down the compute unit to which the workgroup was
bound, and schedule the workgroup for later execution on a different
compute unit.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts
(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
and various engineers from GPU vendors), is that (1) GPU vendors will
not commit to the occupancy-bound execution model that they currently
implement, due to the need for multi-tasking and energy throttling,
yet (2) GPU vendors will not guarantee fair scheduling, due to the
runtime overhead associated with supporting full preemption of
workgroups~\cite{ISCAPAPERSreeMentioned}.  Vendors instead wish to
retain the essence of the occupancy-bound execution model, which is
simple to implement efficiently, and provide support for preemption
only in key special cases.

@@

We propose \emph{cooperative kernels}, an extension to the GPU
programming model that aims to resolve this impasse.  A GPU kernel
that requires workgroups to be fairly scheduled is identified as a
\emph{cooperative} kernel, and is written using two additional
language primitives, $\offerkill$ and $\offerfork$, which must be
placed carefully by the programmer.

At a point where the cooperative kernel could proceed with fewer
workgroups, a workgroup can execute $\offerkill$, offering to
sacrifice itself to the scheduler.  This indicates to the scheduler
that the workgroup would ideally continue with its execution, but that
if required the scheduler may destroy the workgroup for good: the
cooperative kernel is prepared to deal with either scenario.

At a point where the cooperative kernel could benefit from having
additional workgroups join the computation, a workgroup can execute
$\offerfork$.  This indicates to the scheduler that the cooperative
kernel is prepared to proceed with the existing set of workgroups, but
is able to take advantage of having one or more additional workgroups
commence execution directly after the $\offerfork$ program point.

The use of $\offerfork$ and $\offerkill$ creates a contract between
the scheduler and (the programmer of the) cooperative kernel, with
both functional and non-functional requirements.  Functionally, the
scheduler must guarantee that the workgroups executing a cooperative
kernel are fairly scheduled, while the cooperative kernel must be
robust to workgroups leaving and joining the computation in response
to $\offerkill$ and $\offerfork$.  Non-functionally, a cooperative
kernel must ensure that $\offerkill$ is executed frequently enough
that the scheduler can accommodate soft-real time constraints,
e.g.\ allowing a smooth frame-rate for graphics, or allowing compute
units to be powered down quickly when required.  The scheduler should
respond to $\offerkill$ and $\offerfork$ calls in a manner that avoids
under-utilisation of hardware resources by the cooperative kernel,
e.g.\ the scheduler should not kill workgroups unless they are
required for another purpose, and should facilitate forking of
additional workgroups when possible.

@@

\item The model has no impact on the development of regular
  (non-cooperative) kernels: these can be programmed exactly as they
  are now, as can graphics shading routines written using APIs such as
  OpenGL, Vulkan and DirectX.

\item The model is backwards-compatible: a non-blocking algorithm
  implementation that works on today's GPU platforms can be upgraded
  to use our language extensions for cooperative scheduling, and will
  behave exactly as it does at present if the new $\offerkill$ and
  $\offerfork$ functions are simply ignored.

@@

\item Our experiments across four GPUs from two vendors show that our
  portable, proof-of-concept implementation of cooperative kernels can
  provide efficient multi-tasking of cooperative and non-cooperative
  tasks.  A GPU vendor implementation could likely provide even better
  efficiency, by leveraging platform-specific hardware, driver and
  compiler features.

@@

\item \emph{Cooperative kernels}, an extension to the GPU programming model to support the scheduling requirements of blocking algorithms.  We describe the semantics of the programming model in detail (\mysec\ref{sec:cooperativekernels}).

@@

\item An experience report of porting \NumAlgorithms{} irregular GPU
  algorithms to use cooperative kernels, and experimental results
  assessing the performance overhead associated with cooperation and
  the responsiveness to high priority tasks that cooperation enables,
  across four GPUs from two vendors (\mysec\ref{sec:experiments}).

>>>>

Irregular algorithms usually require \emph{blocking synchronization}
between workgroups, to balance load and communicate intermediate
results.  For example, many graph algorithms employ a level-by-level strategy, requiring a global barrier between each level, and 
work stealing algorithms require each workgroup
to maintain a queue, which is typically protected by a mutex to enable
stealing by other workgroups.

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling, which in the context of GPUs means fair scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if one workgroup holds a mutex, an unfair scheduler may cause
another workgroup to spin-wait forever for the mutex to be
released.  Similarly, an unfair scheduler can cause a workgroup to spin-wait
indefinitely at a global barrier so that other workgroups do not reach the barrier.

\paragraph{A degree of fairness: occupancy-bound execution}
GPU schedulers are \emph{not} fair. the current GPU
programming models---OpenCL~\cite{opencl2Spec}, \ADComment{Tyler, please add best CUDA since I've asked you to add it below in any case} CUDA~\cite{...}  and
HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
workgroups, and current GPUs do not
provide fair scheduling in practice.  Roughly speaking, each workgroup
executing a GPU kernel is mapped to a hardware \emph{compute
  unit}.\footnote{In practice the situation can be more complex: if a
  kernel is sufficiently simple and if workgroups are small enough,
  multiple workgroups may be mapped to a single compute unit.  We assume, for the purpose of the current discussion, that the kernel is sufficiently complex that this does not happen.}
%
The simplest way for a GPU driver to handle 
the case where more workgroups are launched than there are compute
units is via an \emph{occupancy-bound}
execution model~\cite{DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{Tyler, please add best cite for OB model; I guess our OOPSLA paper introduced the term, but we should cite that + Owens?}  With this model, once a workgroup has commenced execution on a
compute unit (it has become \emph{occupant}), the workgroup has
exclusive access to the compute unit until it has finished execution,
after which a subsequent workgroup can be bound to the compute unit.
Experiments suggest that the occupancy-bound execution model is widely
implemented by today's GPU devices and drivers~\cite{DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{For the previous cite, should we cite other papers as well as ours?}  It is
evidently a simple model for device vendors to implement efficiently.

The occupancy-bound execution model does not guarantee fair scheduling
between workgroups: if all compute units are occupied then a
not-yet-occupant workgroup will not be scheduled until some occupant
workgroup completes execution.  Yet the execution model \emph{does}
provide fair scheduling between \emph{occupant} workgroups, which are
bound to separate compute units that operate in parallel.  Current GPU
implementations of irregular data parallel algorithms assume the
occupancy-bound execution model, which they exploit by launching 
no more workgroups than there are available compute
units~\cite{...}. \ADComment{Tyler, please pick best cite for previous
  claim.} This works because \emph{today}'s GPUs seem to provide the
occupancy-bound execution model.

@@

Despite its practical prevalence, none of the current GPU programming
models actually mandate occupancy-bound execution.  Further, there are
clear reasons why this model is undesirable.
% from the point-of-view of
%GPU vendors and end users.
First, the execution model does not enable
multi-tasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  If a long-running
kernel occupies all compute units, the GPU cannot be used for other
tasks, such as graphics rendering, in the meantime.
%Precisely because
%of this problem, operating systems employ a GPU \emph{watchdog} to
%heuristically detect and kill long-running computations on the GPU
%being used for display rendering~\cite{DBLP:conf/iwocl/SorensenD16}.
Second, \emph{energy throttling} is
an important concern for battery-powered devices, such as smartphones
and tablets~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to be able to
power down one or more compute units if the battery level is low or if
the user puts the device into a low power state.  Implementing energy
throttling would require being able to suspend the execution of a
workgroup, power down the compute unit to which the workgroup was
bound, and schedule the workgroup for later execution on a different
compute unit.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts
(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
and various engineers from GPU vendors), is that (1) GPU vendors will
not commit to the occupancy-bound execution model that they currently
implement, due to the need for multi-tasking and energy throttling,
yet (2) GPU vendors will not guarantee fair scheduling, due to the
runtime overhead associated with supporting full preemption of
workgroups~\cite{ISCAPAPERSreeMentioned}.  Vendors instead wish to
retain the essence of the occupancy-bound model, which is
simple to implement efficiently, supporting preemption
only in key special cases.

@@

We propose \emph{cooperative kernels}, an extension to the GPU
programming model that aims to resolve this impasse.  A GPU kernel
that requires workgroups to be fairly scheduled is identified as a
\emph{cooperative} kernel, and is written using two additional
language primitives, $\offerkill$ and $\offerfork$, placed by the programmer.

At a point where the cooperative kernel could proceed with fewer
workgroups, a workgroup can execute $\offerkill$, offering to
sacrifice itself to the scheduler.  This indicates that the workgroup
would ideally continue with its execution but that, if required, the
scheduler may destroy the workgroup for good: the cooperative kernel
is prepared to deal with either scenario.

At a point where the cooperative kernel could benefit from having
additional workgroups join the computation, a workgroup can execute
$\offerfork$.  This indicates that the cooperative
kernel is prepared to proceed with the existing set of workgroups, but
is able to take advantage of having one or more additional workgroups
commence execution directly after the $\offerfork$ program point.

The use of $\offerfork$ and $\offerkill$ creates a contract between
the scheduler and (the programmer of the) cooperative kernel.
Functionally, the
scheduler must guarantee that the workgroups executing a cooperative
kernel are fairly scheduled, while the cooperative kernel must be
robust to workgroups leaving and joining the computation in response
to $\offerkill$ and $\offerfork$.  Non-functionally, a cooperative
kernel must ensure that $\offerkill$ is executed frequently enough
that the scheduler can accommodate soft-real time constraints,
e.g.\ allowing a smooth frame-rate for graphics, or allowing compute
units to be powered down quickly when required.  The scheduler should
respond to $\offerkill$ and $\offerfork$ calls in a manner that avoids
under-utilisation of hardware resources by the cooperative kernel,
e.g.\ the scheduler should not kill workgroups unless they are
required for another purpose, and should facilitate forking of
additional workgroups when possible.

@@

\item The model has no impact on the development of regular
  (non-cooperative) compute and graphics kernels: these can be programmed exactly as they
  are now.
%, as can graphics shading routines written using APIs such as
%OpenGL, Vulkan and DirectX.

\item The model is backwards-compatible: if $\offerkill$ and
  $\offerfork$ are ignored, a cooperative blocking kernel will behave
  exactly as a regular blocking kernel does on current GPUs.

@@

\item Our experiments with a range of applications using three GPUs show that the model can enable efficient multi-tasking of cooperative and non-cooperative
  tasks.

@@

\item \emph{Cooperative kernels}, an extension to the GPU programming model to support the scheduling requirements of blocking algorithms (\mysec\ref{sec:cooperativekernels}).

@@

\item An experience report of porting \NumAlgorithms{} irregular GPU
  algorithms to use cooperative kernels, and experimental results
  using three GPUs assessing the overhead associated with cooperative
  kernels, and the responsiveness of the multi-tasking that the
  approach enables (\mysec\ref{sec:experiments}).
